---
title: "Peak 2 gene association example"
output: html_notebook
---

In this notebook, we will show an example implementation for association of accessibility of regulatory peaks to gene expression. 

```{r}
# ## Install required packages
# install.packages(c("tidyr", "dplyr"))
# BiocManager::install(c("SingleCellExperiment", "GenomicRanges", "scran", "tidyr", "Matrix", "ensembldb", 
#                        "EnsDb.Hsapiens.v86", 'BSgenome.Hsapiens.UCSC.hg38'))

library(SingleCellExperiment)
library(GenomicRanges)
library(scran) ## For variable gene selection
library(tidyr)
library(ensembldb)
library(EnsDb.Hsapiens.v86) ## Remember to pick your genome!
library(BSgenome.Hsapiens.UCSC.hg38)
library(dplyr)
```

Define some utils
```{r}
## String - GRanges conversion
## Borrowed from Signac functions 
## https://satijalab.org/signac/reference/GRangesToString.html
StringToGRanges <- function(regions, sep = c("-", "-"), ...) {
  ranges.df <- data.frame(ranges = regions)
  ranges.df <- separate(
    data = ranges.df,
    col = "ranges",
    sep = paste0(sep[[1]], "|", sep[[2]]),
    into = c("chr", "start", "end")
  )
  granges <- makeGRangesFromDataFrame(df = ranges.df, ...)
  return(granges)
}

GRangesToString <- function(grange, sep = c("-", "-")) {
  regions <- paste0(
    as.character(x = seqnames(x = grange)),
    sep[[1]],
    start(x = grange),
    sep[[2]],
    end(x = grange)
  )
  return(regions)
}

# Extend genomicRanges
# 
extend <- function(x, upstream=0, downstream=0)     
{
    if (any(strand(x) == "*"))
        warning("'*' ranges were treated as '+'")
    on_plus <- strand(x) == "+" | strand(x) == "*"
    new_start <- start(x) - ifelse(on_plus, upstream, downstream)
    new_end <- end(x) + ifelse(on_plus, downstream, upstream)
    ranges(x) <- IRanges(new_start, new_end)
    trim(x)
}

```



## Load multiome data

```{r}
atac_sce <- readRDS("./processed_data/Multiome_ATAC_ext_trajectory_SCE.RDS")
rna_sce <- readRDS("./processed_data/Multiome_RNA_ext_trajectory_SCE.RDS")

atac_sce <- atac_sce[rowSums(counts(atac_sce)) != 0,]

```

## Define peak-gene pairs to test 

Here we select highly variable genes (HVG) and for each gene we test for associations in peaks located within 50 kbs of the gene body.  

```{r}
# Find peaks close to features of interest
#
# @param peaks_gr GenomicRanges object containing peaks
# @param features_gr GenomicRanges object containing features (e.g. genes)
# @param d distance to include peak, in bps (default 50000)
# @param feat_anno column in `features_gr@elementMetadata` containing annotation to name features (if NULL converts Granges to string)
#
# @return Sparse adjacency matrix indicating hits (peaks in rows, features in columns)
peak2feature <- function(peaks_gr, features_gr, d=50000, feat_anno=NULL){
  seqlevelsStyle(features_gr) <- seqlevelsStyle(peaks_gr)
  
  ## Find peaks overlapping the search range around the features
  ext_gr <- extend(features_gr, upstream = d, downstream = d)
  ovs <- findOverlaps(peaks_gr, ext_gr)
  
  ## Define identifiers for peaks and features
  all_peaks <- GRangesToString(peaks_gr, sep = c(":", '-'))
  if (is.null(feat_anno)) {
    all_feats <- GRangesToString(features_gr, sep = c(":", '-'))
  } else {
    all_feats <- features_gr@elementMetadata[[feat_anno]]
  }
  
  ## Build adjacency matrix for hits
  adj_mat <- Matrix::Matrix(data=0, nrow = length(all_peaks), ncol=length(all_feats))
  for (i in unique(subjectHits(ovs))) {
    # if (length(adj_mat[queryHits(ovs[subjectHits(ovs)==i]),i]) > 0) {
    adj_mat[queryHits(ovs[subjectHits(ovs)==i]),i] <- 1
    # }
  }
  colnames(adj_mat) <- all_feats
  rownames(adj_mat) <- all_peaks
  
  adj_mat
}

## Select HVGs
dec_rna <- modelGeneVar(rna_sce)
hvg_rna <- getTopHVGs(dec_rna, n=1000)

## Extract genomic locations of these genes
hg38_genes_gr <- genes(EnsDb.Hsapiens.v86)
hvg_gr <- hg38_genes_gr[hvg_rna[hvg_rna %in% names(hg38_genes_gr)]]

## Extract genomic locations of peaks (trying on a subset of peaks)
peaks_gr <- rowRanges(atac_sce)

## Find peaks proximal to HVGs
p2g_mapping <- peak2feature(peaks_gr, hvg_gr, 
             d = 50000 ## within a range of 50k bps
             )

## Filter peaks with at least one overlap
p2g_mapping_filtered <- p2g_mapping[rowSums(p2g_mapping) > 0,]
colnames(p2g_mapping_filtered) <- hvg_rna[hvg_rna %in% names(hg38_genes_gr)]
p2g_mapping_filtered <- as(p2g_mapping_filtered, "dgTMatrix")
```

## Test for associations.

To testing for correlation between gene expression and chromatin accessibility, we implement the following steps:

1. compute a gene-peak correlation coefficient (Spearman) by correlating normalized scATAC-seq peak peak counts with the corresponding gene’s expression across all ATAC-RNA paired cells (observed correlation);
2. For each tested peak, define a set of 30 background peaks with matching total accessibility and GC content to the tested one, using the `getBackgroundPeaks` function implemented in ChromVAR
3. Calculate the correlation between gene expression and the accessibility of all background peaks, to define a null peak-gene correlation distribution that is independent of peak-gene proximity
4. Test for significance with Z-test: we calculate the background mean and standard deviation of spearman correlation. The Z score is calculated by `Z = (cor_observed-cor_bg_mean)/cor_bg_sd`, and converted to a p-value based on the normal distribution.

This code adapts helper functions from `https://github.com/buenrostrolab/stimATAC_analyses_code`

```{r}
run_p2gene_cor <- function(rna_sce, atac_sce, pairnames){
  pairnames <- as.data.frame(pairnames)
  uniq_peaks <- unique(pairnames[,1])
  uniq_genes <- unique(pairnames[,2])
  
  A = counts(atac_sce) # ATAC matrix
  R = counts(rna_sce) # RNA matrix
  
  all_peaks <- GRangesToString(rowRanges(atac_sce), sep = c(":", "-"))
  uniq_peaks_ixs <- which(all_peaks %in% uniq_peaks)
  uniq_genes_ixs <- which(rownames(R) %in% uniq_genes)
  
  M1 <- as.matrix(Matrix::t(A[uniq_peaks_ixs,,drop=FALSE])) 
  M2 <- as.matrix(Matrix::t(R[uniq_genes_ixs,,drop=FALSE])) 
  
  cormat <- cor(x = M1, y = M2, method = "spearman")
  rownames(cormat) <- uniq_peaks
  colnames(cormat) <- uniq_genes
  cors <- sapply(1:nrow(pairnames), function(i) cormat[pairnames[i,1], pairnames[i,2]])
  pairnames <- cbind(pairnames, cors)
  colnames(pairnames) <- c("peak", "gene", "cor")
  return(pairnames)
}

pairnames <- cbind(rownames(p2g_mapping_filtered)[p2g_mapping_filtered@i], colnames(p2g_mapping_filtered)[p2g_mapping_filtered@j])

## Filter to a single gene for testing
g <- unique(pairnames[,2])[1]
pairnames <- pairnames[pairnames[,2]==g,]

## Compute correlation of true peak-gene pairs
cor_observed <- run_p2gene_cor(rna_sce, atac_sce, pairnames)
colnames(cor_observed)[3] <- "cor_observed"

## Compare to accessibility of background peaks
n_iter = 30 ## how many background peaks to use
sognif_alpha = 0.01 ## Significance threshold for p-values

## Find background peaks
atac_chromvar_sce <- addGCBias(atac_sce[uniq_peaks_ixs,], genome=BSgenome.Hsapiens.UCSC.hg38)
rownames(atac_chromvar_sce) <- uniq_peaks
bg <- getBackgroundPeaks(atac_chromvar_sce, niterations = n_iter) ## This can take a while
rownames(bg) <- rownames(atac_chromvar_sce)

## Find null correlation to background
for (i in 1:n_iter){
  pairnames_bg <- pairnames[,1:2] # Initialize gene-peak pairing to observed
  pairnames_bg[,1] <- uniq_peaks[bg[pairnames[,1],i]] # replace with bg peaks for iteration

  cor_bg_i <- run_p2gene_cor(rna_sce, atac_sce, pairnames_bg)
  cor_observed <- cbind(cor_observed, cor_bg_i[,"cor"] )
  colnames(cor_observed)[ncol(cor_observed)] <- paste0("cor_bg_", i)
  }

## Test for significance 
permCols <- 4:(ncol(bg)+3)
cor_observed_filt <- cor_observed[cor_observed[,"cor_observed"] > 0,]   # Filter to positive correlations
  
# Remove multi-mapping peaks (force 1-1 mapping)
cor_observed_filt <- data.frame(cor_observed_filt) %>% group_by(peak) %>% filter(cor_observed==max(cor_observed))

# # Z test pval
cor_observed_filt$rBgSD <- matrixStats::rowSds(as.matrix(cor_observed_filt[,permCols]))
cor_observed_filt$rBgMean <- rowMeans(cor_observed_filt[,permCols])
cor_observed_filt$pvalZ <- 1-stats::pnorm(q = cor_observed_filt$cor_observed, mean = cor_observed_filt$rBgMean,sd = cor_observed_filt$rBgSD)

## Filter to significant results
signif_cor_observed <- cor_observed_filt[cor_observed_filt$pvalZ < signif_alpha, c("peak","gene","cor_observed","pvalZ")]
```

## Save output 

```{r}
## Save peak 2 gene associations
write.csv(signif_cor_observed, "./outputs/signif_peak2genes_mappings.csv")
```

---

```{r}
sessionInfo()
```

